\pkg{shorts}: An R Package for Modeling Short Sprints

### 2020-12-30 paper.tex
Preprint version

\documentclass[
]{jss}

@@

% Pandoc header

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (\textless5s) of maximal efforts \citep{mangineSpeedForcePower2014}, whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60m) \citep{ward-smithEnergyConversionStrategies2001}, but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40m)\citep{brownAssessmentLinearSprinting2004}. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}. Radar and laser technology are frequently used laboratory-grade methods \citep{buchheitMechanicalDeterminantsAcceleration2014, edwardsSprintAccelerationCharacteristics2020, jimenez-reyesRelationshipVerticalHorizontal2018, marcote-pequenoAssociationForceVelocity2019} but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20m), which can now be incorporated into the method for determining sprint mechanical properties \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}. This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance. The \pkg{shorts} package \citep{R-shorts}, written in the R language \citep{R-base}, represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}.

@@

\caption{\label{tab:four-athletes-table}Four athletes with different MSS and MAC parameters.}
\centering
\begin{tabular}[t]{l|r|r|r}
\hline
Athlete & MSS & MAC & TAU\\
\hline
Athlete A & 12 & 10 & 1.20\\
\hline
Athlete B & 12 & 6 & 2.00\\
\hline
Athlete C & 8 & 10 & 0.80\\
\hline
Athlete D & 8 & 6 & 1.33\\
\hline
\end{tabular}
\end{table}

@@

Figure \ref{fig:four-athletes-kinematics} depicts distance, velocity, and acceleration over time (from 0 to 6s).

@@

\caption[Kinematic characteristic of four athletes with different MSS and MAC parameters over a period of 0 to 6seconds]{Kinematic characteristic of four athletes with different MSS and MAC parameters over a period of 0 to 6seconds.}\label{fig:four-athletes-kinematics}
\end{figure}
\end{CodeChunk}

Plotting acceleration against velocity (Figure \ref{fig:four-athletes-profile}), we will get Acceleration-Velocity profile, which is linear, according to the mathematical model.

@@

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40m), (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \eqref{eq:velocity-time} is performed in \pkg{shorts} package using non-linear least squares regression implemented in the \texttt{nls} function in the \pkg{base} R \citep{R-base} and \texttt{nlme} function in the \pkg{nlme} package \citep{R-nlme} for the mixed-effect models.

@@

Let's consider an example of an athlete with MSS equal to 9 \(ms^-1\), TAU equal to 1.3, and MAC equal to 6.92 \(ms^-2\) performing 40m sprint with timing gates positioned at each 10m split. For split times, distance is a predictor, and time is the outcome variable, thus the equation \eqref{eq:velocity-time} becomes:

@@

\(W\) in equation \eqref{eq:time-distance} represents Lambert's W function \citep{R-LambertW}. MSS and TAU parameters are estimated using \texttt{model\_using\_splits} function:

@@

Maximal relative power (PMAX) from the output is estimated using \(\frac{MSS \times MAC}{4}\), which disregards the air resistance. \texttt{time\_correction} and \texttt{distance\_corection} parameters will be covered later in the article.

Besides providing \emph{residual standard error} (RSE), \pkg{shorts} functions provide additional model fit estimators. Additional information can be gained by exploring the returned object, particularly object returned from the \texttt{nls} function:

\begin{CodeChunk}
\begin{CodeInput}
R> summary(m1$model)
\end{CodeInput}
\begin{CodeOutput}
#> 
#> Formula: corrected_time ~ TAU * I(LambertW::W(-exp(1)^(-distance/(MSS * 
#>     TAU) - 1))) + distance/MSS + TAU
#> 
#> Parameters:
#>     Estimate Std. Error t value Pr(>|t|)    
#> MSS   9.0121     0.0155     581  3.0e-06 ***
#> TAU   1.3083     0.0066     198  2.5e-05 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> Residual standard error: 0.00249 on 2 degrees of freedom
#> 
#> Number of iterations to convergence: 4 
#> Achieved convergence tolerance: 3.11e-06
\end{CodeOutput}
\end{CodeChunk}

Once we have estimated MSS and TAU, we can use \texttt{predict\_} family of functions to predict various relationships (i.e.~time at distance, acceleration at distance, velocity at time):

@@

To estimate force production at distance or time (using \texttt{predict\_force\_at\_distance} and \texttt{predict\_force\_at\_time} functions), as well as power production (using \texttt{predict\_power\_at\_distance} and \texttt{predict\_power\_at\_time} functions), one needs to take into account the air resistance. Air resistance (N) is estimated using \texttt{get\_air\_resistance} function, which takes velocity, body mass (kg), body height (m), barometric pressure (Torr), air temperature (\(C^\circ\)), and wind velocity (\(ms^-1\)) as parameters (please refer to \citet{arsacModelingEnergetics100m2002}, \citet{samozinoSimpleMethodMeasuring2016}, and \citet{vaningenschenauCanCyclePower1991} for more information):

@@

When estimating force and power, the air resistance parameters can be set using \texttt{"..."}, which are forwarded to the \texttt{get\_air\_resistance}:

@@

The easiest way to get all kinematics for short sprints is to use \texttt{predict\_kinematics} functions:

\begin{CodeChunk}
\begin{CodeInput}
R> df <- predict_kinematics(
+   m1,
+   max_time = 6,
+   frequency = 100,
+   # Additional parameters forwarded to get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
R> 
R> head(df)
\end{CodeInput}
\begin{CodeOutput}
#>   time distance velocity acceleration air_resistance force power
#> 1 0.00 0.000000   0.0000         6.89        0.07536   551   0.0
#> 2 0.01 0.000344   0.0686         6.84        0.05609   547  37.5
#> 3 0.02 0.001371   0.1367         6.78        0.03978   543  74.2
#> 4 0.03 0.003076   0.2043         6.73        0.02636   539 110.0
#> 5 0.04 0.005455   0.2714         6.68        0.01576   534 145.0
#> 6 0.05 0.008502   0.3379         6.63        0.00792   530 179.2
#>   relative_power
#> 1          0.000
#> 2          0.469
#> 3          0.928
#> 4          1.375
#> 5          1.813
#> 6          2.241
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> require(tidyverse)
R> 
R> df <- pivot_longer(data = df, cols = -2)
R> 
R> ggplot(df, aes(x = distance, y = value)) +
+   theme_bw(8) +
+   facet_wrap(~name, scales = "free_y") +
+   geom_line(alpha = 0.7) +
+   ylab(NULL) +
+   xlab("Distance (m)")
\end{CodeInput}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-7-1} \end{center}

@@

@@

Another valuable addition for sport scientists and coaches is the ability to determine the distances and times where 90\% of maximum sprinting speed is reached, or where peak power is within 90\% range. To identify these values, \pkg{shorts} package comes with \texttt{find\_} family of functions:

@@

Sprint performance is often evaluated with a group of athletes (e.g., soccer club) representing a single strata of interest. Sports scientists can estimate individual profiles, or utilize mixed-effects models. To perform mixed-effects models in \pkg{shorts} for split times, one can use \texttt{mixed\_model\_using\_splits} function. To demonstrate this functionality, we load the \texttt{split\_times} dataset provided in the \pkg{shorts} package:

\begin{CodeChunk}
\begin{CodeInput}
R> data(split_times)
R> 
R> # Mixed model
R> m2 <- mixed_model_using_splits(
+   data = split_times,
+   distance = "distance",
+   time = "time",
+   athlete = "athlete",
+ 
+   # Select random effects
+   # Default is MSS and TAU
+   random = MSS + TAU ~ 1
+ )
R> 
R> m2
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>               8.065               0.655              12.309 
#>                PMAX     time_correction distance_correction 
#>              24.818               0.000               0.000 
#> 
#> Estimated random model parameters
#> ----------------------------------
#>     athlete  MSS   TAU  MAC PMAX time_correction distance_correction
#> 1     James 9.69 0.847 11.4 27.7               0                   0
#> 2       Jim 7.83 0.505 15.5 30.4               0                   0
#> 3      John 7.78 0.727 10.7 20.8               0                   0
#> 4 Kimberley 8.57 0.802 10.7 22.9               0                   0
#> 5  Samantha 6.45 0.395 16.3 26.4               0                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>    0.0260    0.9998   -0.0293    0.0496    0.0496    0.0214    0.0172 
#>      MAPE 
#>    0.9019
\end{CodeOutput}
\end{CodeChunk}

Additional information about mixed-effects model performed using the \texttt{nlme} package \citep{R-nlme} can be obtained using \texttt{summary}:

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> df <- predict_kinematics(m2, max_time = 10)
R> 
R> df <- pivot_longer(df, cols = c(-1, -3))
R> 
R> ggplot(
+   filter(df, distance < 40),
+   aes(x = distance, y = value, group = athlete, color = athlete)
+ ) +
+   theme_bw(8) +
+   facet_wrap(~name, scales = "free_y") +
+   geom_line(alpha = 0.7) +
+   ylab(NULL) +
+   xlab("Distance (m)") +
+   theme(
+     legend.position = "top",
+     legend.title = element_blank())
\end{CodeInput}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-11-1} \end{center}

@@

Let's consider the same example of an athlete with MSS equal to 9\(ms^-1\), TAU equal to 1.3, and MAC equal to 6.92\(ms^-2\) performing 40m sprint with velocity estimated using radar run (in this case with 1Hz sampling rate).

@@

@@

Mixed-effects model using radar data is done using \texttt{mixed\_model\_using\_radar} function. To perform mixed model, let's load data that comes with \pkg{shorts} package.

\begin{CodeChunk}
\begin{CodeInput}
R> data("radar_gun_data")
R> 
R> m4 <- mixed_model_using_radar(
+   radar_gun_data,
+   time = "time",
+   velocity = "velocity",
+   athlete = "athlete"
+ )
R> 
R> m4
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>                8.30                1.01                8.24 
#>                PMAX     time_correction distance_correction 
#>               17.09                0.00                0.00 
#> 
#> Estimated random model parameters
#> ----------------------------------
#>     athlete   MSS   TAU  MAC PMAX time_correction distance_correction
#> 1     James 10.00 1.111 9.00 22.5               0                   0
#> 2       Jim  8.00 0.889 9.00 18.0               0                   0
#> 3      John  8.00 1.069 7.48 15.0               0                   0
#> 4 Kimberley  9.01 1.286 7.01 15.8               0                   0
#> 5  Samantha  6.50 0.685 9.50 15.4               0                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>    0.0516    0.9994   -0.2191    0.1983    0.2191    0.0516    0.0395 
#>      MAPE 
#>       NaN
\end{CodeOutput}
\end{CodeChunk}

@@

One source of error in the modeled estimation using a radar gun is the time synchronization. In theory, synchronization is ideal when a sprint is initiated at \(t=0\) (i.e., \(v(t=0) = 0\)). In practice, this is often not the case. Let's use our athlete and add and deduct 0.5s to simulate an error in synchronization and its effect on estimated MSS and TAU.

@@

@@

As can be seen from the example, TAU is affected by an error in synchronization of time with velocity. The potential solution incorporated into the \pkg{shorts} package involves estimation of the \emph{time correction} parameter using the following equation:

@@

This model is incorporated in the \texttt{model\_using\_radar\_with\_time\_correction} function:

@@

When using \texttt{predict\_} family of functions, one can provide estimated time correction to get predictions at original time scale.

>>>>

% To remove when JSS.cls is fixed
% https://github.com/rstudio/rticles/issues/329
\RequirePackage[2020/02/02]{latexrelease}
\documentclass[
]{jss}

@@

% Pandoc citation processing

% Pandoc header
\usepackage[htt]{hyphenat} \usepackage{booktabs} \usepackage{microtype}

\usepackage{flafter}

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (\textless5 s) of maximal efforts \citep{mangineSpeedForcePower2014}, whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) \citep{ward-smithEnergyConversionStrategies2001}, but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)\citep{brownAssessmentLinearSprinting2004}. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}. Radar and laser technology are frequently used laboratory-grade methods \citep{buchheitMechanicalDeterminantsAcceleration2014, edwardsSprintAccelerationCharacteristics2020, jimenez-reyesRelationshipVerticalHorizontal2018, marcote-pequenoAssociationForceVelocity2019} but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20 m), which can now be incorporated into the method for determining sprint mechanical properties \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}. This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance. The \pkg{shorts} package \citep{R-shorts}, written in the R language \citep{R-base}, represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}.

@@

\caption{\label{tab:four-athletes-table}Four athletes with different MSS and MAC parameters.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Athlete & MSS & MAC & TAU\\
\midrule
Athlete A & 12 & 10 & 1.20\\
Athlete B & 12 & 6 & 2.00\\
Athlete C & 8 & 10 & 0.80\\
Athlete D & 8 & 6 & 1.33\\
\bottomrule
\end{tabular}
\end{table}

@@

Figure \ref{fig:four-athletes-kinematics} depicts distance, velocity, and acceleration over time (from 0 to 6 s).

@@

\caption[Kinematic characteristic of four athletes with different MSS and MAC parameters over a period of 0 to 6 seconds]{Kinematic characteristic of four athletes with different MSS and MAC parameters over a period of 0 to 6 seconds.}\label{fig:four-athletes-kinematics}
\end{figure}
\end{CodeChunk}

Plotting acceleration against velocity (Figure \ref{fig:four-athletes-profile}), we will get \emph{Acceleration-Velocity Profile}, which is linear, according to the mathematical model. If the athlete's body mass (kg) is known, as well as additional air resistance parameters (see \protect\hyperlink{air-resistance-and-the-calculation-of-force-and-mechanical-power}{Air resistance and the calculation of force and mechanical power} section of this paper), \emph{Force-Velocity Profile} can be estimated (see \protect\hyperlink{force-velocity-profile}{Force-Velocity profile} section of this paper).

@@

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40 m), (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \eqref{eq:velocity-time} is performed in \pkg{shorts} package using non-linear least squares regression implemented in the \texttt{nls()} function in the \pkg{base} R \citep{R-base} and \texttt{nlme()} function in the \pkg{nlme} package \citep{R-nlme} for the mixed-effect models.

@@

Let's consider an example of an athlete with MSS equal to 9 \(ms^{-1}\), TAU equal to 1.3, and MAC equal to 6.92 \(ms^{-2}\) performing 40m sprint with timing gates positioned at each 10m split. For split times, distance is a predictor, and time is the outcome variable, thus the equation \eqref{eq:velocity-time} becomes:

@@

\(W\) in equation \eqref{eq:time-distance} represents Lambert's W function \citep{R-LambertW}. MSS and TAU parameters are estimated using \texttt{model\_using\_splits()} function:

@@

Maximal relative power (PMAX) from the output is estimated using \(\frac{MSS \times MAC}{4}\), which disregards the air resistance. \texttt{time\_correction} and \texttt{distance\_corection} parameters will be covered later in the paper.

Besides providing \emph{residual standard error} (RSE), \pkg{shorts} functions provide additional model fit estimators. Additional information can be gained by exploring the returned object, particularly object returned from the \texttt{nls()} function:

\begin{CodeChunk}
\begin{CodeInput}
R> summary(m1)
\end{CodeInput}
\begin{CodeOutput}
#> 
#> Formula: corrected_time ~ TAU * I(LambertW::W(-exp(1)^(-distance/(MSS * 
#>     TAU) - 1))) + distance/MSS + TAU
#> 
#> Parameters:
#>     Estimate Std. Error t value Pr(>|t|)    
#> MSS   9.0121     0.0155     581 0.000003 ***
#> TAU   1.3083     0.0066     198 0.000025 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> Residual standard error: 0.00249 on 2 degrees of freedom
#> 
#> Number of iterations to convergence: 4 
#> Achieved convergence tolerance: 0.00000311
\end{CodeOutput}
\end{CodeChunk}

To extract estimated model parameters, use S3 \texttt{coef()} method:

\begin{CodeChunk}
\begin{CodeInput}
R> coef(m1)
\end{CodeInput}
\begin{CodeOutput}
#>                 MSS                 TAU                 MAC 
#>                9.01                1.31                6.89 
#>                PMAX     time_correction distance_correction 
#>               15.52                0.00                0.00
\end{CodeOutput}
\end{CodeChunk}

To create a simple plot of the model, use S3 \texttt{plot()} method, which returns \pkg{ggplot2} \citep{R-ggplot2} object:

\begin{CodeChunk}
\begin{CodeInput}
R> plot(m1) + theme_bw(8)
\end{CodeInput}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-4-1} \end{center}

Once we have estimated MSS and TAU, we can use \texttt{predict\_XXX()} family of functions to predict various relationships (i.e., time at distance, acceleration at distance, velocity at time, etc.):

@@

To estimate force production at distance or time (using \texttt{predict\_force\_at\_distance()} and \texttt{predict\_force\_at\_time()} functions), as well as power production (using \texttt{predict\_power\_at\_distance()} and \texttt{predict\_power\_at\_time()} functions), one needs to take into account the air resistance. Air resistance (N) is estimated using \texttt{get\_air\_resistance()} function, which takes velocity, body mass (kg), body height (m), barometric pressure (Torr), air temperature (\(C^\circ\)), and wind velocity (\(ms^-1\)) as parameters (please refer to \citet{arsacModelingEnergetics100m2002}, \citet{samozinoSimpleMethodMeasuring2016}, and \citet{vaningenschenauCanCyclePower1991} for more information):

@@

When estimating force and power, the air resistance parameters can be set using \texttt{"..."}, which are forwarded to the \texttt{get\_air\_resistance()}:

@@

The easiest way to get all kinematics and kinetics for short sprints is to use \texttt{predict\_kinematics()} function:

\begin{CodeChunk}
\begin{CodeInput}
R> df <- predict_kinematics(
+   m1,
+   max_time = 6,
+   frequency = 100,
+   # Additional parameters forwarded to get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
R> 
R> head(df)
\end{CodeInput}
\begin{CodeOutput}
#>   time distance velocity acceleration bodymass net_horizontal_force
#> 1 0.00 0.000000   0.0000         6.89       80                  551
#> 2 0.01 0.000344   0.0686         6.84       80                  547
#> 3 0.02 0.001371   0.1367         6.78       80                  543
#> 4 0.03 0.003076   0.2043         6.73       80                  539
#> 5 0.04 0.005455   0.2714         6.68       80                  534
#> 6 0.05 0.008502   0.3379         6.63       80                  530
#>   air_resistance horizontal_force horizontal_force_relative
#> 1        0.07536              551                      6.89
#> 2        0.05609              547                      6.84
#> 3        0.03978              543                      6.78
#> 4        0.02636              539                      6.73
#> 5        0.01576              534                      6.68
#> 6        0.00792              530                      6.63
#>   vertical_force resultant_force resultant_force_relative power
#> 1            785             959                     12.0   0.0
#> 2            785             957                     12.0  37.5
#> 3            785             954                     11.9  74.2
#> 4            785             952                     11.9 110.0
#> 5            785             950                     11.9 145.0
#> 6            785             947                     11.8 179.2
#>   relative_power    RF force_angle
#> 1          0.000 0.575        54.9
#> 2          0.469 0.572        55.1
#> 3          0.928 0.569        55.3
#> 4          1.375 0.566        55.5
#> 5          1.813 0.563        55.7
#> 6          2.241 0.560        55.9
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> require(tidyverse)
R> 
R> variable_names <- colnames(df)
R> 
R> df <- pivot_longer(data = df, cols = -2) %>%
+   mutate(name = factor(name, levels = variable_names))
R> 
R> ggplot(df, aes(x = distance, y = value)) +
+   theme_bw(8) +
+   facet_wrap(~name, scales = "free_y") +
+   geom_line(alpha = 0.7) +
+   ylab(NULL) +
+   xlab("Distance (m)")
\end{CodeInput}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-9-1} \end{center}

@@

These kinematic and kinetic variables are utilized in \protect\hyperlink{force-velocity-profile}{Force-Velocity profile} estimation, which is covered later in this paper.

@@

Another valuable addition for sport scientists and coaches is the ability to determine the distances and times where 90\% of maximum sprinting speed is reached, or where peak power is within 90\% range. To identify these values, \pkg{shorts} package comes with \texttt{find\_XXX()} family of functions:

@@

Sprint performance is often evaluated with a group of athletes (e.g., soccer club) representing a single strata of interest. Sports scientists can estimate individual profiles, or utilize mixed-effects models. To perform mixed-effects models in \pkg{shorts} for split times, one can use \texttt{mixed\_model\_using\_splits()} function. To demonstrate this functionality, we load the \texttt{split\_times} dataset provided in the \pkg{shorts} package:

\begin{CodeChunk}
\begin{CodeInput}
R> data(split_times)
R> 
R> # Mixed model
R> m2 <- mixed_model_using_splits(
+   data = split_times,
+   distance = "distance",
+   time = "time",
+   athlete = "athlete",
+ 
+   # Select random effects
+   # Default is MSS and TAU
+   random = MSS + TAU ~ 1
+ )
R> 
R> m2
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>               8.065               0.655              12.309 
#>                PMAX     time_correction distance_correction 
#>              24.818               0.000               0.000 
#> 
#> Estimated random model parameters
#> ---------------------------------
#>     athlete  MSS   TAU  MAC PMAX time_correction distance_correction
#> 1     James 9.69 0.847 11.4 27.7               0                   0
#> 2       Jim 7.83 0.505 15.5 30.4               0                   0
#> 3      John 7.78 0.727 10.7 20.8               0                   0
#> 4 Kimberley 8.57 0.802 10.7 22.9               0                   0
#> 5  Samantha 6.45 0.395 16.3 26.4               0                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>    0.0260    0.9998   -0.0293    0.0496    0.0496    0.0214    0.0172 
#>      MAPE 
#>    0.9019
\end{CodeOutput}
\end{CodeChunk}

Additional information about mixed-effects model performed using the \texttt{nlme} package \citep{R-nlme} can be obtained using \texttt{summary()} function:

@@

S3 method \texttt{coef()} when applied on mixed-model result will return both the fixed and random effects:

\begin{CodeChunk}
\begin{CodeInput}
R> coef(m2)
\end{CodeInput}
\begin{CodeOutput}
#> $fixed
#>                 MSS                 TAU                 MAC 
#>               8.065               0.655              12.309 
#>                PMAX     time_correction distance_correction 
#>              24.818               0.000               0.000 
#> 
#> $random
#>     athlete  MSS   TAU  MAC PMAX time_correction distance_correction
#> 1     James 9.69 0.847 11.4 27.7               0                   0
#> 2       Jim 7.83 0.505 15.5 30.4               0                   0
#> 3      John 7.78 0.727 10.7 20.8               0                   0
#> 4 Kimberley 8.57 0.802 10.7 22.9               0                   0
#> 5  Samantha 6.45 0.395 16.3 26.4               0                   0
\end{CodeOutput}
\end{CodeChunk}

To create a simple plot of the model, use S3 \texttt{plot()} method:

\begin{CodeChunk}
\begin{CodeInput}
R> plot(m2) + theme_bw(8)
\end{CodeInput}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-14-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> df <- predict_kinematics(m2, max_time = 10)
R> 
R> variable_names <- colnames(df)
R> 
R> df <- pivot_longer(df, cols = c(-1, -3)) %>%
+   mutate(name = factor(name, levels = variable_names))
R> 
R> ggplot(
+   filter(df, distance < 40),
+   aes(x = distance, y = value, group = athlete, color = athlete)
+ ) +
+   theme_bw(8) +
+   facet_wrap(~name, scales = "free_y") +
+   geom_line(alpha = 0.7) +
+   ylab(NULL) +
+   xlab("Distance (m)") +
+   theme(
+     legend.position = "top",
+     legend.title = element_blank())
\end{CodeInput}

@@

Let's consider the same example of an athlete with MSS equal to 9 \(ms^{-1}\), TAU equal to 1.3, and MAC equal to 6.92 \(ms^{-2}\) performing 40m sprint with velocity estimated using radar run (in this case with 1 Hz sampling rate).

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot(m3) + theme_bw(8)
\end{CodeInput}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-17-1} \end{center}

@@

Mixed-effects model using radar data is done using \texttt{mixed\_model\_using\_radar()} function. To perform mixed model, let's load data that comes with \pkg{shorts} package.

\begin{CodeChunk}
\begin{CodeInput}
R> data("radar_gun_data")
R> 
R> m4 <- mixed_model_using_radar(
+   radar_gun_data,
+   time = "time",
+   velocity = "velocity",
+   athlete = "athlete"
+ )
R> 
R> m4
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>                8.30                1.01                8.24 
#>                PMAX     time_correction distance_correction 
#>               17.09                0.00                0.00 
#> 
#> Estimated random model parameters
#> ---------------------------------
#>     athlete   MSS   TAU  MAC PMAX time_correction distance_correction
#> 1     James 10.00 1.111 9.00 22.5               0                   0
#> 2       Jim  8.00 0.889 9.00 18.0               0                   0
#> 3      John  8.00 1.069 7.48 15.0               0                   0
#> 4 Kimberley  9.01 1.286 7.01 15.8               0                   0
#> 5  Samantha  6.50 0.685 9.50 15.4               0                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>    0.0516    0.9994   -0.2191    0.1983    0.2191    0.0516    0.0395 
#>      MAPE 
#>       NaN
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> plot(m4) + theme_bw(8)
\end{CodeInput}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-20-1} \end{center}

\hypertarget{force-velocity-profile}{%
\subsection{Force-Velocity profile}\label{force-velocity-profile}}

To create \emph{Force-Velocity Profile} (FVP) using single athlete estimated sprint model parameters (i.e., TAU and MSS), you can use \texttt{get\_FV\_profile()} function. When estimating FVP, athlete body mass (kg) can be set using \texttt{bodymass} parameter, while the air resistance parameters can be set using \texttt{"..."}, which are forwarded to the \texttt{get\_air\_resistance()} function. Details of the FVP method implemented in the \pkg{shorts} package, as well as the interpretation from a sprint training perspective, are covered elsewhere \citep{haugenPowerForceVelocityProfilingSprinting2020, morinInterpretingPowerForceVelocityProfiles2016, morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}.

\begin{CodeChunk}
\begin{CodeInput}
R> # To create Force-Velocity Profile
R> fvp <- get_FV_profile(
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU,
+   bodymass = 80,
+   # Additional parameters forwarded to get_air_resistance
+   # Otherwise, defaults are used
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
R> 
R> fvp
\end{CodeInput}
\begin{CodeOutput}
#> Estimated Force-Velocity Profile
#> --------------------------------
#>      bodymass            F0        F0_rel            V0          Pmax 
#>      80.00000     544.51032       6.80638       9.36184    1274.40402 
#> Pmax_relative      FV_slope  RFmax_cutoff         RFmax           Drf 
#>      15.93005      -0.72703       0.30000       0.48779      -0.06676 
#>        RSE_FV       RSE_Drf 
#>       1.54814       0.00447
\end{CodeOutput}
\end{CodeChunk}

To plot FVP kinematics and kinetics (which are exactly the same as generated by the \texttt{predict\_kinematics()} function), use S3 \texttt{plot()} function. By default, FVP estimated kinetics are plotted against velocity (on x-axis).

\begin{CodeChunk}
\begin{CodeInput}
R> plot(fvp) + theme_bw(8)
\end{CodeInput}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-22-1} \end{center}

To plot FVP estimated kinetics against time, use \texttt{type\ =\ "time"} parameter:

\begin{CodeChunk}
\begin{CodeInput}
R> plot(fvp, "time") + theme_bw(8)
\end{CodeInput}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-23-1} \end{center}

@@

One source of error in the modeled estimation using a radar gun is the time synchronization. In theory, synchronization is ideal when a sprint is initiated at \(t=0\) (i.e., \(v(t=0) = 0\)). In practice, this is often not the case. Let's use our athlete and add and deduct 0.5 s to simulate an error in synchronization and its effect on estimated MSS and TAU.

@@

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-24-1} \end{center}

@@

As can be seen from the example, all estimated parameters are affected by an error in synchronization of time with velocity (with MSS being the least affected in this example). The potential solution incorporated into the \pkg{shorts} package involves estimation of the \emph{time correction} parameter using the following equation:

@@

This model is incorporated in the \texttt{model\_using\_radar\_with\_time\_correction()} function:

@@

When using \texttt{predict\_XXX()} family of functions, one can provide estimated time correction to get predictions at original time scale.

\begin{CodeChunk}
\begin{CodeInput}
R> # Using the true time
R> predict_velocity_at_time(
+   time = df$`true time`,
+   MSS = m5$parameters$MSS,
+   TAU = m5$parameters$TAU
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 0.00 4.83 7.07 8.11 8.59 8.81 8.91
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # Using time with sync issues
R> predict_velocity_at_time(
+   time = df$`0.5s added`,
+   MSS = m8$parameters$MSS,
+   TAU = m8$parameters$TAU,
+   time_correction = m8$parameters$time_correction
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 0.0000782 4.8299729 7.0681475 8.1053182 8.5859434 8.8086652
#> [7] 8.9118746
\end{CodeOutput}
\end{CodeChunk}
